Refunctionalization of Abstract Abstract Machines (Functional Pearl)

### 2018-07-07 paper/paper.tex
small fixes + grant numbers

  In this paper, we explain their functional correspondence and demonstrate how to
  syntactically transform small-step abstract abstract machines into big-step abstract
  definitional interpreters.
  Building on known semantic interderivation techniques from the concrete evaluation
  setting, the transformations include linearization, lightweight fusion, disentanglement,
  refunctionalization, and the left inverse of the CPS transform.
  Linearization expresses nondeterministic choice through first-order data types,
  after which refunctionalization transforms the first-order data types that represent continuations
  into higher-order functions.
  The refunctionalized AAM is an abstract interpreter written in continuation-passing style (CPS)
  with two layers of continuations, which can be converted back to direct style with delimited
  control operators. Based on the known correspondence between delimited control and monads,
  we demonstrate that the explicit use of monads in abstract definitional interpreters
  is optional.

@@

\Citet{van2012systematic,van2010abstracting} proposed the Abstracting abstract machines (AAM)
methodology which provides a recipe for constructing sound abstract interpreters
for higher-order functional languages from concrete abstract machines.
%~\cite{van2012systematic, van2010abstracting}.
Given a concrete small-step abstract machine that models a store
(e.g., the CESK machine, Krivine's machine, etc.), if we allocate continuations
in the store and draw all allocated addresses (both for regular values and
for continuations) from a finite set, then
we obtain an abstract interpreter with a finite state space, which can be used for
performing sound static analysis and is guaranteed to terminate.
By using different address allocators \cite{Gilray:2016:ACP:2951913.2951936},
one can further instantiate different polyvariant control-flow analyses.

Applying the same idea to monadic definitional interpreters, \citet{darais2017abstracting}
built monadic abstract definitional interpreters (ADIs) in big-step style.
One of the advantages of a monadic interpreter is
that it is modular and composable. By changing the underlying monads, the definition
of the interpreter is not modified, but we can recover different semantics, including
the concrete semantics and various abstract semantics such as context-sensitivity and
abstract garbage collection \cite{Sergey:2013:MAI:2491956.2491979}.

Broadly speaking, abstract abstract machines and abstract definitional interpreters are
different forms of abstract interpreters. They are obtained by applying a combination
of abstractions to their concrete counterparts: abstract machines and definitional
interpreters, respectively. An interesting question, and the subject of this paper, is
how one can derive one abstract semantic artifact given its counterpart.
The main contribution of this paper is to explain the functional correspondence
between these abstract semantic artifacts, and to demonstrate the concrete
syntactic transformation steps necessary to transform an abstract abstract machine
into its corresponding abstract definitional interpreter. The reverse direction
is comparatively easier, and can be obtained by following the respective inverse
steps in the opposite order. This contribution fills an intellectual gap in our
understanding of abstract semantic artifacts, and it also opens up further practical
avenues for constructing static analysis tools based on principled and well-understood
techniques.

@@

In ANF, an expression is either an atomic expression or a @let@ expression.
A restriction exists which states that all function applications must be administered
within a @let@ expression and then bound to a variable name under the current environment.
Both the operator and operand of function applications are atomic expressions.
An atomic expression $ae$ is either a variable or a literal @lambda@ term, either of which
can be evaluated in a single step. We also assume that all the variable names in the program
 are unique.

The abstract syntax in Scala is shown as follows. We assume that the source program conforms
to the ANF convention, and as such, do not enforce it in the term structure of Scala constructs.

@@

\subsection{Abstracting Abstract Machines} \label{aam}
Abstracting abstract machines (AAM) is a systematic methodology that derives sound
abstract interpreters for higher-order functional languages from concrete abstract
machines \cite{van2012systematic, van2010abstracting}. An abstract abstract machine
implements computable abstract semantics which approximates the run-time behaviors of
programs.
Since the state space of concrete execution is possibly infinite, the key insight of the
AAM approach when analyzing programs is to allocate both variable bindings and continuations
on the store, and then bound the addresses space to be finite. Since each component
of the machine state is finite, the abstracted machine-state space is also finite,
and therefore computable.

@@

Similar to the CESK machine, the machine state of an AAM has a control, an environment, a store,
and continuation, as well as a timestamp. However, there are several notable differences
between an AAM's store and the CESK machine's store. In AAM, the store maps addresses to sets of
values; it stores all possible values for a particular address. As such, dereferencing
addresses becomes nondeterministic. Also, the store performs \emph{joining}, rather than
overwriting, when updating elements. Furthermore, the continuations are likewise allocated
on the store instead of formed into a linked list, and the continuation component becomes
an address that maps to a set of continuations in the store instead of directly embedded
in the state.

@@

  This work was supported in part by the
  \grantsponsor{GS100000001}{National Science
    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
  No.~\grantnum{GS100000001}{mmmmmmm} and by the
  \grantsponsor{GS100000001}{DOE}{XXX} under Grant
  No.~\grantnum{GS100000001}{nnnnnnn}.  \note{TODO: grants}
\end{acks}

>>>>

  In this paper, we explain their functional correspondence and demonstrate how to
  systematically transform small-step abstract abstract machines into big-step abstract
  definitional interpreters.
  Building on known semantic interderivation techniques from the concrete evaluation
  setting, the transformations include linearization, lightweight fusion, disentanglement,
  refunctionalization, and the left inverse of the CPS transform.
  Linearization expresses nondeterministic choice through first-order data types,
  after which refunctionalization transforms the first-order data types that represent continuations
  into higher-order functions.
  The refunctionalized AAM is an abstract interpreter written in continuation-passing style (CPS)
  with two layers of continuations, which can be converted back to direct style with delimited
  control operators. Based on the known correspondence between delimited control and monads,
  we demonstrate that the explicit use of monads in abstract definitional interpreters
  is optional.

@@

\Citet{van2012systematic,van2010abstracting} proposed the abstracting abstract machines (AAM)
methodology which provides a recipe for constructing sound abstract interpreters
for higher-order functional languages from concrete abstract machines.
%~\cite{van2012systematic, van2010abstracting}.
Given a concrete small-step abstract machine that models a store
(e.g., the CESK machine, Krivine's machine, etc.), if we allocate continuations
in the store and draw all allocated addresses (both for regular values and
for continuations) from a finite set, then
we obtain an abstract interpreter with a finite state space, which can be used for
performing sound static analysis and is guaranteed to terminate.
By using different address allocators \cite{Gilray:2016:ACP:2951913.2951936},
one can further instantiate different polyvariant control-flow analyses.

Applying the same idea to monadic definitional interpreters, \citet{darais2017abstracting}
showed how to build monadic abstract definitional interpreters (ADIs) in big-step style.
One of the advantages of a monadic interpreter is
that it is modular and composable. By changing the underlying monads, the definition
of the interpreter is not modified, but we can recover different semantics, including
the concrete semantics and various abstract semantics such as context-sensitivity and
abstract garbage collection \cite{Sergey:2013:MAI:2491956.2491979}.

Broadly speaking, abstract abstract machines and abstract definitional interpreters are
different forms of abstract interpreters. They are obtained by applying a combination
of abstractions to their concrete counterparts: abstract machines and definitional
interpreters, respectively. An interesting question, and the subject of this paper, is
how one can derive one abstract semantic artifact given the other.
The main contribution of this paper is to explain the functional correspondence
between these abstract semantic artifacts, and to demonstrate the concrete
syntactic transformation steps necessary to transform an abstract abstract machine
into its corresponding abstract definitional interpreter. The reverse direction
is comparatively easier, and can be obtained by following the respective inverse
steps in the opposite order. This contribution fills an intellectual gap in our
understanding of abstract semantic artifacts, and it also opens up further practical
avenues for constructing static analysis tools based on principled and well-understood
techniques.

@@

In ANF, an expression is either an atomic expression or a @let@ expression.
A restriction exists which states that all function applications (and only those) must be administered
within a @let@ expression and then bound to a variable name under the current environment.
Both the operator and operand of function applications are atomic expressions.
An atomic expression $ae$ is either a variable or a literal @lambda@ term, either of which
can be evaluated in a single step. We also assume that all the variable names in the program
 are unique.

The abstract syntax is represented in Scala as follows. We assume that the source program conforms
to the ANF convention, and as such, do not enforce it in the term structure of Scala constructs.

@@

\subsection{Abstracting Abstract Machines} \label{aam}
Abstracting abstract machines (AAM) is a systematic methodology that derives sound
abstract interpreters for higher-order functional languages from concrete abstract
machines \cite{van2012systematic, van2010abstracting}. An abstract abstract machine
implements a computable abstract semantics which approximates the run-time behaviors of
programs.
Since the state space of concrete execution is possibly infinite, the key insight of the
AAM approach when analyzing programs is to allocate both variable bindings and continuations
on the store, and then bound the addresses space to be finite. Since each component
of the machine state is finite, the abstracted machine-state space is also finite,
and therefore computable.

@@

Similar to the concrete CESK machine, the machine state of the derived AAM has a control component, an environment, a store,
a continuation, as well as a timestamp. However, there are several notable differences
between an AAM's store and the CESK machine's store. In AAM, the store maps addresses to sets of
values; it stores all possible values for a particular address. As such, dereferencing
addresses becomes nondeterministic. Also, the store performs \emph{joining}, rather than
overwriting, when updating elements. Furthermore, the continuations are likewise allocated
on the store instead of formed into a linked list, and the continuation component becomes
an address that maps to a set of continuations in the store instead of directly embedded
in the state.

@@

  This work was supported in part by 
  NSF awards 1553471 and 1564207, and DOE award DE-SC0018050.
  % \grantsponsor{GS100000001}{National Science
  %   Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
  % No.~\grantnum{GS100000001}{nnnnnnn} and Grant
  % No.~\grantnum{GS100000001}{mmmmmmm} and by the
  % \grantsponsor{GS100000001}{DOE}{XXX} under Grant
  % No.~\grantnum{GS100000001}{nnnnnnn}.  \note{TODO: grants}
\end{acks}
